E-Graphs With Bindings

### 2025-01-26 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
made formatting more compact by inlining some maths

We build on the technology of string diagrams, which is by now well-developed in a number of areas. 
String diagrams have the advantage of being a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, which are a fundamental data structure in algorithm design.
These categorical techniques provide insight in how to solve the problems of binding and capture-avoiding substitution in e-graphs, as we develop in this paper.

@@

Revisiting the simple earlier example $\lambda f . f ((\lambda x . f x) 2)$, in Figure~\ref{fig:de-brujin-string}, observe that the two occurrences of $f$ are now shared.
$\lambda$-abstraction is represented by a rounded box with the bound variable a dangling wire attached to the frame,
the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=lambda_unit_string] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}, 
and sharing a \emph{contraction node} 
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

@@

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the \textit{directed acyclic graph} (DAG) representation of terms to include equivalence classes of subterms (or subgraphs).
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Figure~\ref{fig:e-graph-example}, where it is shown how a series of rewrite rules can be non-destructively applied to an e-graph~\cite{EggPaper}.

@@

First, let us compare how substitution works in conventional e-graphs as compared to our proposed approach.
In Figure~\ref{fig:e-graph-substitution} we show successive iterations of a conventional e-graph for $(\lambda x . (y + y) + x) 1$ following $\beta$-reductions.
A subterm for an explicit substitution must be introduced, which is deemed equivalent to the top-level application term (hence they are inside the same dashed box), to be rewritten using the rules for explicit substitution at the top of each column.
At the end, all explicit substitution nodes can be removed.
The same example is then given again in Figure~\ref{fig:e-string-substitution}, this time using string diagrams.

>>>>

We build on the technology of string diagrams, by now well-developed in a number of areas. 
String diagrams are a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, which are a fundamental data structure in algorithm design.
These categorical techniques provide insight in how to solve the problems of binding and capture-avoiding substitution in e-graphs, as we develop in this paper.

@@

Revisiting the simple earlier example $\lambda f . f ((\lambda x . f x) 2)$, in Fig.~\ref{fig:de-brujin-string}, observe that the two occurrences of $f$ are now shared.
$\lambda$-abstraction is represented by a rounded box with the bound variable a dangling wire attached to the frame,
the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=lambda_unit_string] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}, 
and sharing a \emph{contraction node} 
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

@@

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the \textit{directed acyclic graph} (DAG) representation of terms to include equivalence classes of subterms (or subgraphs).
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Fig.~\ref{fig:e-graph-example}, where it is shown how a series of rewrite rules can be non-destructively applied to an e-graph~\cite{EggPaper}.

@@

First, let us compare how substitution works in conventional e-graphs as compared to our proposed approach.
In Fig.~\ref{fig:e-graph-substitution} we show successive iterations of a conventional e-graph for $(\lambda x . (y + y) + x) 1$ following $\beta$-reductions.
A subterm for an explicit substitution must be introduced, which is deemed equivalent to the top-level application term (hence they are inside the same dashed box), to be rewritten using the rules for explicit substitution at the top of each column.
At the end, all explicit substitution nodes can be removed.
The same example is then given again in Fig.~\ref{fig:e-string-substitution}, this time using string diagrams.

### 2025-01-26 LICS_2025/e_graphs_with_bindings/sections/categorical_semantics.tex
fixes
made formatting more compact by inlining some maths

\begin{definition}
A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar
\[
\tau \coloneqq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2}
\],
where $\phi$ ranges over the set of operators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

\begin{definition}[Semilattice-enriched category~\cite{Borceux_1994}]
\label{def:slat-cat}
    A semilattice-enriched category $\mathbb{C}$ is defined by the following data:
    \begin{itemize}
        \item a set of objects $\mathbb{C}$
        \item for every pair of objects $A,B \in \mathbb{C}$ --- a hom-object (or, hom-semilattice) $\mathbb{C}(A,B) \in \catname{SLat}$
        \item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism
        \[
            \circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)
        \]
        \item for every object $A \in C$ --- a unit morphism
        \[
        u_{A} : I \to \mathbb{C}(A,A), \text{ where $I$ is the monoidal unit for $\catname{SLat}$}
        \]
    \end{itemize}
    such that particular coherence diagrams commute.
\end{definition}
As the composition is a semilattice homomorphism (morphism of $\catname{SLat}$), we have the following equations 
\[
(f + g); h = f;h + g;h \quad f;(g+h) = f;g + f;h
\]

@@

\begin{definition}[Semilattice-enriched SMC~\cite{enriched_monoidal}]\label{def:enriched-prop}
    A \textit{semilattice enriched strict SMC}  $\mathbb{C}$ is given by a semilattice-enriched category $\mathbb{C}$ as above that additionally has
    \begin{itemize}
    \item a unit object $I_{C} \in \mathbb{C}$
    \item for every pair of objects $A,B \in \mathbb{C}$ --- an object $A \otimes B \in \mathbb{C}$
    \item for all $A,B,C,D$ --- a tensor morphism $ - \otimes_{C} - : \mathbb{C}(A,C) \otimes \mathbb{C}(B,D) \to \mathbb{C}(A \otimes B, C \otimes D)$
    \end{itemize}
    such that particular coherence diagrams commute.
    \end{definition}
    The latter morphism being in $\catname{SLat}$ implies the following equations for $\otimes$:
    \[
    f \otimes (g+h) = f \otimes g + f \otimes h \qquad (f+g) \otimes h = f \otimes h + g \otimes h
    \]

    Again, we take $\otimes$ to bind more tightly than $+$.

@@

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994})
        There is a 2-adjunction 
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
        % \[\begin{tikzcd}
        % 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        \mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
        \vspace{-1mm}
        \]
        that is induced by a usual free-forgetful adjunction

>>>>

\begin{definition}
A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar:
\[
\tau \coloneqq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2},
\]
where $\phi$ ranges over the set of operators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

\begin{definition}[Semilattice-enriched category~\cite{Borceux_1994}]
\label{def:slat-cat}
    A semilattice-enriched category $\mathbb{C}$ is defined by the following data:
    \begin{itemize}
        \item a set of objects $\mathbb{C}$
        \item for every pair of objects $A,B \in \mathbb{C}$ --- a hom-object (or, hom-semilattice) $\mathbb{C}(A,B) \in \catname{SLat}$
        \item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism 
        $\circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)$
        \item for every object $A \in C$ --- a unit morphism 
        $u_{A} : I \to \mathbb{C}(A,A)$, where $I$ is the monoidal unit for $\catname{SLat}$.
    \end{itemize}
    such that particular coherence diagrams commute.
\end{definition}
As the composition is a semilattice homomorphism (morphism of $\catname{SLat}$), we require that 
$
(f + g); h = f;h + g;h$ and $f;(g+h) = f;g + f;h
$.

@@

\begin{definition}[Semilattice-enriched SMC~\cite{enriched_monoidal}]\label{def:enriched-prop}
    A \textit{semilattice enriched strict SMC}  $\mathbb{C}$ is given by a semilattice-enriched category $\mathbb{C}$ as above that additionally has
    \begin{itemize}
    \item a unit object $I_{C} \in \mathbb{C}$
    \item for every pair of objects $A,B \in \mathbb{C}$ --- an object $A \otimes B \in \mathbb{C}$
    \item for all $A,B,C,D$ --- a tensor morphism $ - \otimes_{C} - : \mathbb{C}(A,C) \otimes \mathbb{C}(B,D) \to \mathbb{C}(A \otimes B, C \otimes D)$
    \end{itemize}
    such that particular coherence diagrams commute.
    \end{definition}
    The latter morphism being in $\catname{SLat}$ implies that for $\otimes$ we have 
    $
    f \otimes (g+h) = f \otimes g + f \otimes h $ and $ (f+g) \otimes h = f \otimes h + g \otimes h
    $.
    As before, we take $\otimes$ to bind more tightly than $+$.

@@

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994})
        There is a 2-adjunction 
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
        % \[\begin{tikzcd}
        % 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        \mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
        \vspace{-1mm}
        \]
        that is induced by the usual free-forgetful adjunction
